---
title: Political Ideology & Front-line House Democrats
date: '2020-03-16'
slug: political-ideologies-of-front-line-house-democrats
output:
  blogdown::html_page:
    toc: true
    toc_depth: 2
tags:
  - rstats  
  - 116th Congress
  - open-government
  - VoteView
bibliography: biblio.bib
link-citations: yes
banner: banners/front_ideo.png
description: 'A super quick investigation into the voting behavior of front-line House Democrats.'
---


## Briefly

A quick look at the voting behavior of the 30 House Democrats that represent congressional districts carried by Trump in 2016.  Using `Rvoteview`.  For a more in-depth account of the characteristics of front-line House Democrats in the 116th Congress, see [this post](https://www.jtimm.net/2019/12/13/frontline-democrats/).



## Front-line House Democrats

I have been going on about the House of Representatives quite a bit lately, especially Democratic members representing Trump districts.  These guys were instrumental in Democrats recapturing the House majority in 2018.  Over two-thirds of these front-liners are freshman members, and as a group they are super vulnerable heading into November 2020.  

I have posted a list of the 30 front-liners as a simple csv, cached as a part of the `uspoliticalextras` data package.  It is available at the link below. 

```{r message=FALSE, warning=FALSE}
library(tidyverse)
url1 <- 'https://raw.githubusercontent.com/jaytimm/uspoliticalextras/master/clean-data-sets/thirty-one-house-democrats.csv'
fl <- read.csv(url(url1)) 
```



## Ideologies in the 116th

So, using the `Rvoteview` (!) package, we obtain DW-Nominate scores for all members in the 116th House.  This session is still in progress, so these numbers will change depending on when they are accessed.  

```{r}
x116 <- Rvoteview::member_search (chamber = 'House', 
                                  congress = 116) %>%
  
  mutate(label = gsub(', .*$', '', bioname),
         party_code = ifelse(bioname %in% fl$house_rep, 
                             'xx', party_code),
         party_name = ifelse(bioname %in% fl$house_rep, 
                             'Frontline Dems', 'Other Dems'))
```


**The plot below** summarizes voting behaviors as approximated by DW-Nominate scores in two dimensions.  Here, our focus is on the first dimension (ie, the x-axis).  The 30 front-liners are marked in orange.  In the aggregate, then, they vote more moderately than their non-front-line Democrat peers.  

```{r fig.height=6}
p <- x116 %>%
  ggplot(aes(x=nominate.dim1, 
             y=nominate.dim2,
             label = label
             )) +
          annotate("path",
               x=cos(seq(0,2*pi,length.out=300)),
               y=sin(seq(0,2*pi,length.out=300)),
               color='gray',
               size = .25) +
  geom_point(aes(color = as.factor(party_code)), 
             size= 2.5, 
             shape= 17) +
  theme_bw() +
  ggthemes::scale_color_stata() +
  theme(legend.position = 'none') +
  labs(title="DW-Nominate ideology scores for the 116th US House",
       subtitle = '30 front-line House Democrats in orange')

p
```



## Focusing on Democrats

Next, we home in a bit on House Democrats.   To add some context to the above plot, we calculate quartiles for DW-Nominate scores among Democrats.  These are summarized in table below, ranging from progressive to moderate.

```{r message=FALSE, warning=FALSE}
dems <- x116 %>%
  filter(party_code %in% c('xx', '100')) 


qq <- data.frame(x = quantile(dems$nominate.dim1, probs = seq(0, 1, 0.25)),
                 stringsAsFactors = FALSE)

qq %>% knitr::kable()
```


We add these quartiles to the plot below, and label front-line House Democrats.  Again, front-liners cluster as a group in terms of roll call voting behavior.  The most notable exception to this pattern is Lauren Underhood (IL-14). She won her district by five points in 2018, and Trump won the district by 4 points in 2016.  It would appear, then, that her voting behavior and the political ideology of her constituents do not especially rhyme.  In  other words, **she represents a Trump district and votes like a progressive**. 

```{r fig.height=6, message=FALSE, warning=FALSE}
p1 <- p +
  xlim(-1, 0) +
  geom_vline(xintercept = qq$x, linetype = 2, color = 'gray') +
  ggrepel::geom_text_repel(
   data  = filter(x116, 
                  bioname %in% fl$house_rep),
   nudge_y =  -0.005,
   direction = "y",
   hjust = 0, 
   size = 2.5)

p1
```


**The table below** summarizes counts of Democrats by front-line status & ideology quartile.  So, roughly 3/4 of front-liners vote in the most moderate Democratic quartile in the House. And all but Underwood are in top 50%.  

```{r}
dems1 <- dems %>%
  mutate(qt = ntile(nominate.dim1, 4))

dems1 %>% 
  group_by(party_name, qt) %>%
  count() %>%
  group_by(party_name) %>%
  mutate(per = round(n/sum(n)*100, 1)) %>%
  knitr::kable(booktabs = T, format = "html") %>%
  kableExtra::kable_styling() %>%
  kableExtra::row_spec(3,  
                       background = "#e4eef4")
```


## Summary

Support this group of House members!! Follow them on Twitter!

```{r echo=FALSE, message=FALSE, warning=FALSE}
leg_dets <- 'https://theunitedstates.io/congress-legislators/legislators-current.csv'

x <- read.csv((url(leg_dets)),
                       stringsAsFactors = FALSE) %>%
  filter(type == 'rep') %>% # & twitter!=''
  mutate(twitter1 = paste0('[', twitter, ']', '(',
                           'https://twitter.com/',
                           twitter, ')'),
         twitter2 = paste0('@', twitter)) %>%
  rename (state_abbrev = state,
          district_code = district) %>%
  inner_join(fl) %>%
  filter(twitter != '')

x %>%
  summarise(twitter1 = paste0(twitter1, collapse = ' | ')) %>%
  knitr::kable()
```



