---
title: 'Twitter in collage: a year in the life of a freshman Congresswoman'
author: ''
date: '2020-01-01'
slug: 2020-01-01a-year-in-twitter-photos
tags:
  - rstats
  - 116th Congress
  - open-government
  - Twitter
banner: banners/collage.png
description: 'Some methods for a year in collage.'
---

## Introduction

A super simple post that summarizes R-based methods for visual summary & collage-building using image attachments on Twitter.   In the process, a bit of a photo homage to Congresswoman Xochitl Torres Small in her first year representing New Mexico's 2nd district.  

```{r message=FALSE, warning=FALSE}
if (!require("pacman")) install.packages("pacman")
pacman::p_load(tidyverse, rtweet, tigris) 
options(tigris_use_cache = TRUE, tigris_class = "sf")
```


```{r message=FALSE, warning=FALSE, include=FALSE}
rtweet::create_token(
  app = "jtimm_twitter",
  consumer_key = "CpTkgrdkRWQTrYSR4EEW4YSo6",
  consumer_secret = "39F4DCY2WtMMqo1lDQMXXCEWrgqvXgRZYE99m4OW8oUuuevzD7",
  access_token = "965072442954690560-H04g14OTNfZMzC7nUvvpvSGB1Cr2ERm",
  access_secret = "Gn2hNKmcLIaXSZxKrh5AfeyttbrXKEeAAnrw8koGHRP3p")
```



## New Mexico's 2nd District

The 2nd congressional district of New Mexico is a super fun district.  It is not my district, but I have a few stomping grounds that way.  Faywood is home to natural hot springs and a STAR-GAZING CHAIR!!!.  Weed has an absolutely lovely frisbee golf course.  Ruidoso has all things: including two disc golf courses and a horse track.

Folks in the district supported Trump in 2016 by a fair margin (+10.2%) and subsequently sent a freshman Democrat to the House in 2018.  Only the second time the district has sent a Democrat to the House in the last 30 years.  Also, it is one of only [five districts](https://www.jtimm.net/2019/12/13/frontline-democrats/#the-5-house-democrats-that-should-probably-vote-against-impeachment) that supported Trump by more than ten points, supported McCain in 2008 & Romney in 2012, and sent a Democrat to Congress in 2018. I have written some previously about [Torres Small's win in 2018](https://www.jtimm.net/2019/01/17/xochitl-torres-small-win-2018/) and the demographics of the district.   

So, a complicated & ideologically diverse district.  And Congresswoman Torres Small does an amazing job representing this diversity. She is one of the few House Dems that engages with Fox News, eg, writing [op-eds](https://www.foxnews.com/opinion/rep-xochitl-torres-small-a-one-size-fits-all-approach-to-border-security-wont-work-heres-what-will) and doing [interviews](https://video.foxnews.com/v/6002988864001/#sp=show-clips).  And she is a [super positive Twitter follow](https://twitter.com/RepTorresSmall) if you are interested in feeling good about the folks that represent us in Congress.

The district is also geographically vast -- and super-rural. Per the table below, NM-02 is the fifth largest district in the country -- only the big rural states with at-large representation are bigger.  So, lots of ground to cover.


```{r message=FALSE, warning=FALSE}
cds <- tigris::congressional_districts(cb = TRUE)

cds %>%
  data.frame() %>%
  arrange(desc(ALAND)) %>%
  slice(1:5) %>%
  mutate(ALAND = round(ALAND/ 2.59e+6,0), # SQUARE MILES
         ALAND = format(ALAND,
                        big.mark=",",
                        scientific=FALSE),
         geo = c('Alaska', 'Montana',
                 'Wyoming', 'South Dakota',
                 'New Mexico - 02')) %>%
  select(geo, GEOID, ALAND)%>%
  knitr::kable()
```


```{r message=FALSE, warning=FALSE, include=FALSE}
xochitl_tweets <- rtweet::get_timeline( 
  "RepTorresSmall", 
  n = 1500,
  check=FALSE) %>%
  mutate(created_at = as.Date(gsub(' .*$', '', created_at))) %>%
  filter(is_quote == 'FALSE' & 
           is_retweet == 'FALSE' & 
           created_at > '2019-01-02' &
           display_text_width > 0)
```


**One hundred days** into her term in the 116th Congress, Congresswoman Torres Small tweeted: 

---

`r xochitl_tweets$text[xochitl_tweets$status_id == '1116748060359524352']` 

---

**So, she is on the move**.  The map below highlights the district in geographical context.  Lots of big districts in the Southwest.  NM-02 is roughly bounded by Mexico, West Texas, ABQ/Santa Fe metros, and the Navajo Nation.  A bit of cultural crossroads, as it were.

```{r fig.height=5}
world <- rnaturalearth::ne_countries(scale = "medium", returnclass = "sf") %>%
  filter(gu_a3 %in% c('USA', 'MEX'))
states <- tigris::states(cb = TRUE)

cds %>%
  mutate(color = ifelse(GEOID == '3502', 'blue', 'gray')) %>%
  ggplot() + 
  geom_sf(data = world,  color = 'darkgray', alpha = .75,  
          fill = '#dae2ba', size = 1.1) + 
  geom_sf(aes(fill = color), color = 'darkgray') + 
  scale_fill_manual(values = c('lightblue', 'gray')) +
  ggsflabel::lims_bbox(cds %>%
                         filter(STATEFP %in% c('04', '48', '35', '32'))) +
  geom_sf(data = states,  color = 'darkgray', alpha = 0,  size = 1.1) + 
  theme(legend.position = 'none') +
  ggtitle('NM-02 in context')
```





## Collaging the year's happenings

So, the goal here is to provide a visual summary (ie, collage) of Congresswoman Torres Small's year representing the nation's fifth largest congressional district using photo attachments on Twitter.  Via the `rtweet` package,  we collect all of `@RepTorresSmall` tweets since she took office at the beginning of 2019.    


```{r eval=FALSE, message=FALSE, warning=FALSE}
xochitl_tweets <- rtweet::get_timeline( 
  "RepTorresSmall", 
  n = 1500,
  check=FALSE) %>%
  mutate(created_at = as.Date(gsub(' .*$', '', created_at))) %>%
  filter(is_quote == 'FALSE' & 
           is_retweet == 'FALSE' & 
           created_at > '2019-01-02' &
           display_text_width > 0)
```


Next, we identify tweets containing photo attachments.  And then download these photos locally.  The code presented here has been modified directly from [this post](https://masalmon.eu/2017/03/19/facesofr/).  For a more detailed walk through of methods, I would recommend having a look.  

```{r message=FALSE, warning=FALSE, include=FALSE}
local_pics <- '/home/jtimm/Desktop/pics' 
home <- '/home/jtimm/jt_work/GitHub/presently'
```


```{r eval=FALSE, message=FALSE, warning=FALSE}
pics <- xochitl_tweets %>%
  filter(!is.na(media_url)) %>%
  select(media_url, created_at)

setwd(local_pics)
lapply(pics$media_url, function (y) {
  magick::image_read(y) %>%
    magick::image_scale("1000") %>%
    magick::image_border('white', '10x10') %>%
    magick::image_write(gsub('^.*/', '', y)) #%>%
    #magick::image_annotate(pics$created_at[y], font = 'Times', size = 50)
  })
```


Next, we shuffle the photos some, and then stack photos as a collection of single column collages.  Again, these intermediary files are stored locally.   

```{r}
files <- dir(local_pics, full.names = TRUE)
set.seed(11)
files <- sample(files, length(files))
```


```{r include=FALSE}
local_cols <- '/home/jtimm/Desktop/cols/cols1/'
```

```{r fig.height=9, message=FALSE, warning=FALSE}
files1 <- files[1:49]
no_rows <- 7
no_cols <- 7

make_column <- function(i, files, no_rows){
  magick::image_read(files[(i*no_rows+1):((i+1)*no_rows)]) %>%
    magick::image_append(stack = TRUE) %>%
    magick::image_write(paste0("cols", i, ".jpg"))}

setwd(local_cols)
walk(0:(no_cols-1), 
     make_column, 
     files = files1,
     no_rows = no_rows)
```


Lastly, we piece together the column collages as a single collage.  For good measure, I created three collages comprised of 7 x 7 = 49 photos.  **A busy year for the Congrtesswoman**.  Tirelessly representing the folks in NM-02 it would seem.  

```{r}
magick::image_read(dir(local_cols, full.names = TRUE)) %>% 
  magick::image_scale("500x2000") %>%
  magick::image_append(stack = FALSE) 
```

```{r echo=FALSE, fig.height=8, message=FALSE, warning=FALSE}
files2 <- files[50:98]

setwd('/home/jtimm/Desktop/cols/cols2')
walk(0:(no_cols-1), 
     make_column, 
     files = files2,
     no_rows = no_rows)
```

```{r echo=FALSE, fig.height=9, message=FALSE, warning=FALSE}
magick::image_read(dir("/home/jtimm/Desktop/cols/cols2", full.names = TRUE)) %>% 
  magick::image_scale("500x2000") %>%
  magick::image_append(stack = FALSE) 
```

```{r echo=FALSE, fig.height=8, message=FALSE, warning=FALSE}
files3 <- files[99:147]

setwd('/home/jtimm/Desktop/cols/cols3')
walk(0:(no_cols-1), 
     make_column, 
     files = files3,
     no_rows = no_rows)
```

```{r echo=FALSE, fig.height=9, message=FALSE, warning=FALSE}
magick::image_read(dir("/home/jtimm/Desktop/cols/cols3", full.names = TRUE)) %>% 
  magick::image_scale("500x2000") %>%
  magick::image_append(stack = FALSE) #%>%
  #magick::image_annotate('@TorresSmall', font = 'Times', size = 100)
```

