---
title: 'time-lining the Trump presidency'
author: ''
date: '2020-11-06'
slug: time-lining-trump-presidency
categories: []
tags: []
description: 'A structured, Wikipedia-based timeline of the Trump presidency.'
banner: banners/timeline.png
---


### Introduction

The [uspols](https://github.com/jaytimm/uspols) package includes a simple function for scraping a Wikipedia-based [timeline of the Trump presidency](https://en.wikipedia.org/wiki/Timeline_of_the_Donald_Trump_presidency) --  `uspols_wiki_timeline()`.  


```{r message=FALSE, warning=FALSE}
library(tidyverse)
```


```{r eval=FALSE}
library(devtools)
devtools::install_github("jaytimm/uspols")
library(uspols) 
```


The function returns an up-to-date table of daily administration happenings from 20 Jan 2017 onward.  Here we walk through a simple work-flow for adding timeline data to time series data. While not an official accounting, *a super quick/easy way to get one's bearings (in a maelstrom) and contextualize data points**.  

```{r}
longs <- uspols::uspols_wiki_timeline()
```


**Table structure is detailed some below**.  The folks at Wikipedia have delineated events (chronologically?) via bullet points per each day of Trump's presidency, which have been enumerated here in the `bullets` column as independent rows.  I have mostly done this to make text easier to read/access/etc.  I am not sure how meaningful event distinctions (for a given day) actually are.   

```{r}
longs %>%
  select(-quarter:-daypres, -dow) %>%
  head() %>%
  DT::datatable(rownames = FALSE, 
                options = list(dom = 't',
                               pageLength = 6,
                               scrollX = TRUE))
```




### A simple use-case

```{r}
summary <- longs %>%
  filter(!is.na(Events)) %>%
  mutate(tokens = tokenizers::count_words(Events)) %>%
  group_by(date) %>%
  mutate(daily_count = sum(tokens)) %>%
  slice(1) %>%
  ungroup()
```



For demonstration, we calculate total word counts of event descriptions per day from the time-line table.  The plot below, then, summarizes these daily counts since inauguration.  The plot itself is fairly meaningless, but the hover-action should be useful. For clarity purposes, only the first event for each day is included in the pop-up,.

```{r message=FALSE, warning=FALSE}
dp <- summary %>%
  mutate(text = stringr::str_wrap(string = Events,
                                  width = 20,
                                  indent = 1,
                                  exdent = 1)) %>%
  
  plotly::plot_ly(x = ~date, 
                  y = ~daily_count,
                  text = ~text,
                  type = 'scatter',
                  mode = 'lines') %>%
  
  plotly::layout(#atitle = "Top 10 Drug Types", 
                 tooltip = c('Events'),
                 yaxis = list (title = "Daily event word count per Wikipedia"))
```




### Example plot with Trump daily event on hover

```{r echo=FALSE, message=FALSE, warning=FALSE,  out.width='100%'}
dp
```


### FIN

For [a more complete perspective](https://github.com/jaytimm/A-timeline-of-the-Trump-presidency#lest-we-forget) ::
