---
title: 'the seven living generations in america' 
date: '2020-06-10'
slug: american-generations
output:
  blogdown::html_page:
    toc: true
    toc_depth: 2
tags:
  - rstats
  - census
bibliography: biblio.bib
link-citations: yes
banner: banners/generations.png
description: "Some methods & a roll call."
---


## American generations

A quick look at the composition of American generations.  Per Pew Research [definitions](https://www.pewresearch.org/fact-tank/2019/01/17/where-millennials-end-and-generation-z-begins/) & US Census data.  In the process, a static perspective on the changing demographics of America.


```{r message=FALSE, warning=FALSE}
library(tidyverse)
gen <- c('Post-Z', 'Gen Z', 'Millennial', 
         'Gen X', 'Boomers', 'Silent', 
         'Greatest')

range <- c('> 2012', '1997-2012', '1981-1996', 
           '1965-1980', '1946-1964', '1928-1945', 
           '< 1927')

gen_desc <- data.frame(rank = 7:1,
                       gen = gen,
                       range = range,
                       stringsAsFactors = FALSE) %>%
  arrange(rank)
```



**A summary:**

```{r echo=FALSE}
DT::datatable(gen_desc, rownames = FALSE, options = list(dom = 't',
                                                         scrollX = TRUE))
```



Four of America's seven living generations are more or less "complete," and only getting smaller (albeit at different rates): `Greatest`, `Silent`, `Boomers`, and `Gen X`. The generation comprised of `Millenials` is complete as well, in that it has been delineated chronologically; however, the group likely continues to grow via immigration.  


While `Gen Z` has been tentatively stamped chronologically by the folks at Pew Research, only the very eldest in the group have just entered the work force. So lot's can happen still. And although we include them here, the `Post-Z` generation is mostly but a thought; half of the group has yet to be born.


## Monthly US population estimates 

*Monthly Postcensal Resident Population plus Armed Forces Overseas*, December 2018.  Made available by the US Census [here](https://www.census.gov/newsroom/press-kits/2019/detailed-estimates.html).  The census has transitioned to a new online interface, and (seemingly) many data sets have been discontinued. Hence, the data set utilized here is slightly dated.


```{r message=FALSE, warning=FALSE}
pops <- read.csv (
  url('https://www2.census.gov/programs-surveys/popest/datasets/2010-2018/national/asrh/nc-est2018-alldata-p-File18.csv')) %>%
  filter(MONTH == '12' & YEAR == '2018') %>%
  gather(key = 'race', value = 'pop', -UNIVERSE:-AGE)
```


  
A more detailed description of the population estimates can be found [here](https://www2.census.gov/programs-surveys/popest/technical-documentation/file-layouts/2010-2018/nc-est2018-alldata.pdf). Note: Race categories reflect non-Hispanic populations.


```{r message=FALSE, warning=FALSE}
race <- c('NHWA', 'NHBA', 'NHIA', 
          'NHAA', 'NHNA', 'NHTOM', 'H')

race1 <- c('White Alone',
           'Black Alone',
           'American Indian Alone',
           'Asian Alone',
           'Native Hawaiian Alone',
           'Two or More Races',
           'Hispanic')

labels <- data.frame(race = race, 
                     race1=race1, 
                     stringsAsFactors = FALSE)

search <- paste(paste0('^',race, '_'), collapse =  '|')
```


The following table details **a random sample of the data set** -- with Pew Research defined generations & estimated year-of-birth.

```{r message=FALSE, warning=FALSE}
gen_pops <- pops %>%
  filter(grepl(search, race)) %>%
  mutate(race = gsub('_.*$', '', race)) %>%
  group_by(AGE, race) %>%
  summarise(pop = sum(pop))%>%
  left_join(labels) %>%
  filter(AGE != '999') %>%
  mutate(yob = 2019 - AGE)  %>% ##
  mutate (gen = case_when (
    yob < 2013 & yob > 1996 ~ 'Gen Z',
    yob < 1997 & yob > 1980 ~ 'Millennial',
    yob < 1981 & yob > 1964 ~ 'Gen X',
    yob < 1965 & yob > 1945 ~ 'Boomers',
    yob < 1946 & yob > 1927 ~ 'Silent',
    yob < 1928 ~ 'Greatest',
    yob > 2012 ~ 'Post-Z')) %>%
  left_join(gen_desc) %>%
  ungroup() %>%
  select(gen, rank, range, race, 
         race1, yob, AGE, pop)

set.seed(999)
gen_pops %>% sample_n(7)  %>%
  select(gen, range, race1:pop) %>%
  DT::datatable(rownames = FALSE, options = list(dom = 't',
                                                 scrollX = TRUE))
```



## Composition of American generations

### Population by generation

The figure below summarizes the US population by generation.  These numbers will vary some depending on the data source.  **Millenials constitute the plurality of Americans**, more recently overtaking a Boomer generation on the wane.

```{r fig.height=5}
gen_pops %>%
  group_by(gen, rank) %>%
  summarize(pop = sum(pop)) %>%
  mutate(lab = round(pop/1000000, 1)) %>%
  ggplot(aes(x = reorder(gen, rank), 
             y = pop, 
             fill = gen)) +
  geom_col(show.legend = FALSE, 
           alpha = 0.75)  +
  geom_text(aes(label = lab), 
            size = 3.5)+
  theme(axis.text.x=element_blank(),
        axis.ticks.x=element_blank())+
  xlab('') + ylab('') +
  coord_flip()+
  ggthemes::scale_fill_stata() +
  theme_minimal() +
  labs(title = 'Population by American generation',
       caption = 'SOURCE: US Census, Monthly Postcensal Resident Population plus Armed Forces Overseas, December 2018.')
```



### Population by single year of age & generation

```{r}
gg <- gen_pops %>% 
  group_by(yob, AGE, gen) %>%
  summarize(tot = sum(pop)) %>%
  group_by(gen) %>%
  mutate(tot = max(tot)) %>% #For labels below.
  filter(yob %in% c('1919', '1928', '1946', '1965', 
                    '1981', '1997', '2013'))
```

The figure below illustrates the US population by single year of age, ranging from the population aged less than a year to the population over 100 (as of December 2018).   Generation membership per single year of age is specified by color.


```{r fig.height=7}
gen_pops %>%
  ggplot(aes(x = AGE, 
             y = pop, 
             fill = gen)) +
  geom_vline(xintercept = gg$AGE,
             linetype =2, 
             color = 'gray', 
             size = .25)+
  
  geom_col(show.legend = FALSE, 
           alpha = 0.85,
           width = .7)   +
  annotate(geom="text", 
           x = gg$AGE - 4.5, 
           y = gg$tot + 70000, 
           label = gg$gen,
           size = 3.25) +
  xlab('Age')+ 
  ylab('') +
  theme_minimal() +
  theme(legend.position="bottom",
        legend.title = element_blank(),
        panel.grid.major.x=element_blank(),
        panel.grid.minor.x=element_blank(),
        panel.grid.minor.y=element_blank()) +
  ggthemes::scale_fill_stata()+
  scale_x_reverse(breaks = rev(gg$AGE)) +
  labs(title = 'American population by single-year age & generation')
```




### Population by single year of age, race & generation

```{r}
gen_pal <- c('#b0bcc1','#ead8c3', '#437193', 
             '#c66767', '#55752f', '#dae2ba', 
             '#7b9bb3')
```


Next, we crosscut the single year of age counts presented above by race & ethnicity.

```{r fig.height=7.25}
gen_pops %>%
  ggplot(aes(x = AGE, 
             y = pop, 
             fill = race1)) +
  geom_area(stat = "identity",
            color = 'white',
            alpha = 0.85) +
  scale_fill_manual(values = gen_pal) +
  geom_vline(xintercept = gg$AGE,
             linetype =2, color = 'gray', size = .25)+
  annotate(geom="text", 
           x = gg$AGE - 4.5, 
           y = gg$tot + 70000, 
           label = gg$gen,
           size = 3.25) +
  xlab('')+ ylab('') +
  theme_minimal() +
  theme(legend.position="bottom",
        legend.title = element_blank(),
        panel.grid.major.x=element_blank(),
        panel.grid.minor.x=element_blank(),
        panel.grid.minor.y=element_blank()) +
  
  scale_x_reverse(breaks = rev(gg$AGE) )+
  labs(title ='American population by age, race & generation')
```



### White America on the wane 

```{r}
white_label <- gen_pops %>% 
  group_by(gen, AGE) %>%
  mutate(per = pop/sum(pop))%>%
  filter(race1 == 'White Alone') %>%
  group_by(gen) %>%
  mutate(per = max(per)) %>% #For labels below.
  arrange(yob) %>%
  filter(yob %in% c('1919', '1928', '1946', '1965', 
                    '1981', '1997', '2013'))
```



The last figure illustrates a **proportional perspective of race & ethnicity in America** by single year of age.   Per figure, generational differences (at a single point in time) can shed light on (the direction of) potential changes in the overall composition of a given populace.  As well as a view of what that populace may have looked like in the past.

```{r fig.height=7.25}
gen_pops %>%
  group_by(gen, AGE) %>%
  mutate(per = pop/sum(pop)) %>%
  ggplot(aes(x = (AGE), 
             y = per, 
             fill = race1)) +
  geom_area(stat = "identity",
            color = 'white',
            alpha = 0.85) +
  geom_hline(yintercept = .5, 
             linetype = 4,
             color = 'white') +
  scale_fill_manual(values = gen_pal) +
  geom_vline(xintercept = gg$AGE,
             linetype = 2, 
             color = 'gray', 
             size = .25)+
  annotate(geom="text", 
           x = gg$AGE-4.5, 
           y = white_label$per - .05, 
           label = gg$gen,
           size = 3.25) +
  xlab('')+ ylab('') +
  theme_minimal() +
  theme(legend.position="bottom",
        legend.title = element_blank(),
        panel.grid.major.x=element_blank(),
        panel.grid.minor.x=element_blank()) +
  
  scale_x_reverse(breaks = rev(gg$AGE)) +
  labs(title = 'American population by age, race & generation')
```



## American generations in (apparent) time & space

 Aggregate race-ethnicity profiles for America's seven living generations are presented in the table below.  Per [American Community Survey (ACS) 2018 5-year estimates](https://github.com/jaytimm/tech-notes-american-politics/blob/master/notes/America_american.md), the `Gen X` race-ethnicity profile (or distribution) is most representative of American demographics overall.  


```{r}
sums <- gen_pops %>%
  group_by(rank, gen, race1) %>%
  summarize(pop = sum(pop))%>%
  group_by(gen) %>%
  mutate(per = round(pop / sum(pop), 3))
```


```{r echo=FALSE}
sums %>%
  select(-pop) %>%
  mutate(per = round(per * 100, 2)) %>%
  spread(race1, per) %>%
  select(-`Native Hawaiian Alone`, 
         -`American Indian Alone`) %>%
  DT::datatable(rownames = FALSE, options = list(dom = 't',
                                                 scrollX = TRUE))
```



```{r message=FALSE, warning=FALSE, include=FALSE}
tidycensus::census_api_key("b508704c99f3ae9bc5b5e7c41e3dd77e59d52722")
#Sys.getenv("CENSUS_API_KEY")
```



### Race-ethnicity profiles for US counties

Here, we compare race-ethnicity profiles for US counties to those of American generations.  Using the `tidycensus` R package, we first obtain county-level race-ethnicity estimates (ACS 2018 5-year);  

```{r message=FALSE, warning=FALSE}
#x <- tidycensus::load_variables(year = 2018, dataset = 'acs5/profile')
variable <- c('DP05_0071P', 'DP05_0077P', 'DP05_0078P',
              'DP05_0079P', 'DP05_0080P', 
              'DP05_0081P', 'DP05_0082P', 
              'DP05_0083P')

variable_name <- c('Hispanic', 'White Alone', 'Black Alone',
                   'American Indian Alone', 'Asian Alone',
                   'Native Hawaiian Alone', 'Some Other Race Alone',
                   'Two or More Races')

gen <-  tidycensus::get_acs(geography = 'county',
                            variables = variable,
                            year = 2018,
                            survey = 'acs5') %>%
  left_join(data.frame(variable, variable_name, stringsAsFactors = FALSE))%>%
  select(-variable, -moe) %>%
  rename(variable = variable_name) %>%
  select(GEOID:NAME, variable, estimate)
```


then, via the `tigris` library, we obtain shapefiles for (1) all US counties and (2) US states. 

```{r message=FALSE, warning=FALSE}
nonx <- c('78', '69', '66', '72', '60', '15', '02')

library(tigris)
options(tigris_use_cache = TRUE, 
        tigris_class = "sf")

us_counties <- tigris::counties(cb = TRUE) %>% 
  filter(!STATEFP %in% nonx)
us_states <- tigris::states(cb = TRUE) %>% 
  filter(!STATEFP %in% nonx)

laea <- sf::st_crs("+proj=laea +lat_0=30 +lon_0=-95")
us_counties <- sf::st_transform(us_counties, laea)
us_states <- sf::st_transform(us_states, laea)
```


### Comparing distributions

The table below highlights the race-ethnicity distribution for my hometown, **Dutchess County, NY**.  GEOID = 36027.


```{r include=FALSE}
dc <- gen %>%
  filter(GEOID == 36027) %>%
  # select(-pop) %>%
  # mutate(per = round(per * 100, 2)) %>%
  spread(variable, estimate) %>%
  select(-`Native Hawaiian Alone`, 
         -`American Indian Alone`,
         -`Some Other Race Alone`,
         -GEOID, -NAME) 
```


```{r echo=FALSE}
dc %>%
  DT::datatable(rownames = FALSE, options = list(dom = 't',
                                                 scrollX = TRUE))
```

<br>

```{r}
which1 <- gen %>%
  mutate(estimate = ifelse(estimate == 0, .0000001, estimate)) %>%
  inner_join(sums, by = c('variable' = 'race1')) %>%
  
  mutate(div = estimate/100 * log((estimate/100)/per)) %>%
  group_by(GEOID, NAME, gen) %>%
  summarize (relative_entropy = round(sum(div), 4)) %>%
  ungroup()
```


Via `Kullback–Leibler divergence` (ie, [relative entropy]((https://github.com/jaytimm/tech-notes-american-politics/blob/master/notes/America_american.md#calculating-relative-entropy))), we compare the demographic profile of Dutchess County to the demographic profiles of each American generation.  Per the table below, the Dutchess County profile is most similar to that of the `Boomer` generation.  *Per the lowest relative entropy value*.  This is not to say that there are more Boomers in Dutchess County; instead, **the racial demographics of Dutchess County are most akin to an America when Boomers were in their prime**.


```{r message=FALSE, warning=FALSE}
which1 %>%
  filter(GEOID == 36027) %>%
  left_join(gen_desc) %>%
  select(rank, NAME, gen, relative_entropy) %>%
  arrange(rank) %>%
  DT::datatable(rownames = FALSE, 
                options = list(dom = 't',
                               scrollX = TRUE,
                               
      rowCallback = DT::JS('function(row, data, index, rowId) {',
                       'console.log(rowId)','if(rowId == 2 ) {',
                       'row.style.backgroundColor = "#dae2ba";','}','}'))) 
```


<br>
```{r}
which2 <- which1 %>%
  group_by(GEOID, NAME) %>%
  filter(relative_entropy == min(relative_entropy)) %>%
  ungroup() 
```


So, while `Gen X` is most representative of America in the aggregate, **per the map below**, the demographics of America's youngest & eldest generations are most prevalent at the county-level.  `Post-Z` in the West, and the `Greatest` in the East & Midwest.  (With `Boomer` & `Millenial` demographics most prevalent in the South.)  New & Old Americas, perhaps.


```{r fig.height=6, message=FALSE, warning=FALSE}
us_counties %>%
  left_join(which2, by = 'GEOID') %>%
  ggplot() + 
  geom_sf(aes(fill = gen),  lwd = 0) + 
  ggthemes::scale_fill_stata() +
  geom_sf(data=us_states, 
          fill = NA, 
          show.legend = F, 
          color="darkgray", 
          lwd=.5) +
  theme(axis.title.x=element_blank(),
        axis.text.x=element_blank(),
        axis.title.y=element_blank(),
        axis.text.y=element_blank(),
        legend.position = 'bottom') +
  labs(title = "Counties & Generations in America")
```



## Summary 

Some different perspectives on the composition of America & American generations, as well as a novel (static) take on the changing demographics of America. 