---
title: 'Modeling roll call voting behavior in the House of Representatives'
date: '2020-02-11'
slug: modeling-roll-call-voting-behavior
output:
  blogdown::html_page:
    toc: true
    toc_depth: 2
tags:
  - rstats
  - 116th Congress
  - open-government
  - VoteView
bibliography: biblio.bib
link-citations: yes
banner: banners/model.png
description: 'A brief go at modeling roll call voting behavior in the US House of Representatives using (1) constituency demographics, (2) house member party affiliation, and (3) house member characteristics. A bit of an homage to McCarty, Poole, and Rosenthal (2016)'
---

## Thoughts initial

A brief (and copycat) go at **modeling roll call voting behavior** in the US House of Representatives using (1) constituency demographics, (2) house member party affiliation, and (3) house member characteristics. This post is based directly on work presented in @mccarty2016polarized, specifically chapter 2 ('Polarized politicians').  Here, we make **transparent & reproducible** some methods using R & open source data sets -- with some fairly comparable results.  

(Also -- just an FYI:  I have started posting super short [snippets of R code on Git Hub](https://github.com/jaytimm/some_technical_notes), mostly related to things text & politics & both. Meant to be quick & useful.)



## Data sets

The basic model we consider here is ~formalized as follows:

* Political ideology~i~ = constituency demographics~i~ + member party~i~ + member characteristics~i~ 

**Can we predict how house members vote** based solely on the characteristics of their constituents, their party affiliation, and their age/gender/race?  Variables comprising these four model components (for district ***i*** and the representative of district ***i***) are summarized in the table below. Variables included here mostly align with those presented in @mccarty2016polarized.  We will describe the details of variables & sources in the following sections.

```{r echo=FALSE, message=FALSE, warning=FALSE}
library(tidyverse)
group <- c('political ideology', 
           'constituency demographics',
           'constituency demographics',
           'constituency demographics',
           'constituency demographics',
           'constituency demographics',
           'member party',
           'member characteristics',
           'member characteristics',
           'member characteristics')

var <- c('dw_NOMINATE_DIM_1', 
         'cd_BACHELORS', 
         'cd_HH_INCOME', 
         'cd_BLACK', 
         'cd_HISPANIC', 
         'cd_IS_SOUTH', 
         'member_PARTY', 
         'member_GENDER', 
         'member_ETHNICITY', 
         'member_AGE')

source <- c('Rvoteview', 
            'ACS 2018 1-Year estimates', 
            'ACS 2018 1-Year estimates',
            'ACS 2018 1-Year estimates', 
            'ACS 2018 1-Year estimates',
            'Dixie + KE + OK', 
            'Rvoteview', 
            'CivilServiceUSA', 
            'CivilServiceUSA',
            'CivilServiceUSA')

meta <- data.frame(group, var, source,
                   stringsAsFactors = FALSE) 
meta %>% knitr::kable()
```



### +Political ideology

The **dependent variable** in the model is roll call voting behavior, or political ideology.  DW-NOMINATE scores from the [VoteView](https://voteview.com/) project are accessed here via the `Rvoteview` [package](https://github.com/voteview/Rvoteview).  Our focus here, then, is first dimension scores from the 116th House.  Scores range from -1 (the most liberal) to +1 (the most conservative).  A score of 0, then, would represent a more moderate voter.

```{r}
nominates <- Rvoteview:: member_search(chamber= 'House', congress = 116) %>%
  select(state_abbrev, district_code, nominate.dim1)
```




### +Member party & member characteristics

The [CivilServiceUSA](https://github.com/CivilServiceUSA) makes available a collection of characteristics about house members, including **age, gender, and race/ethnicity**.  I have cached these data in a package dubbed `uspoliticalextras`, available [here](https://github.com/jaytimm/uspoliticalextras).

```{r eval=FALSE}
devtools::install_github("jaytimm/uspoliticalextras")
```


```{r}
reps <- uspoliticalextras::uspol_csusa_house_bios %>%
  filter(congress == 116) %>%
  select(state_fips:district_code, party, last_name, gender, ethnicity, date_of_birth) %>%
  mutate(age = round(lubridate::interval(date_of_birth, 
                                         Sys.Date())/lubridate::duration(num = 1, 
                                                                         units = "years"))) %>%
  select(-date_of_birth)%>%
  mutate(district_code = ifelse(district_code == 0, 1, district_code),
         ethnicity = ifelse(grepl('middle|multi|nativ|pacific', ethnicity), 
                            'other-race', 
                            ethnicity),
         ethnicity = gsub('-american', '', ethnicity),
         ethnicity = gsub('african', 'black', ethnicity))
```



### +Constituency demographics

Some demographic characteristics for US congressional districts are also included `uspoliticalextras`.  These were accessed via `tidycensus`, and included in the package out of convenience.  **2018 1-year ACS estimates for 12 variables**, which include:

```{r}
unique(uspoliticalextras::uspol_dems2018_house$variable)
```

For modeling purposes, constituency demographics are characterized in terms of % population that is Black, % population that is Hispanic, and % population that has obtained a bachelor's degree or higher.  Median household income for the district is also considered.  Lastly, each district is identified as being a part of the south or not, where southern states are defined as the eleven states of the Confederacy, plus Oklahoma & Kentucky.

```{r}
south <- c('SC', 'MS', 'FL', 
           'AL', 'GA', 'LA', 'TX', 
           'VA', 'AR', 'NC', 'TE',
           'OK', 'KE')

dems <- uspoliticalextras::uspol_dems2018_house %>% 
  spread(variable, estimate) %>%
  mutate(district_code = ifelse(district_code == 0, 1, district_code),
         is_south = ifelse(state_abbrev %in% south, 'Yes', 'No'))
```


We then join the three data sets, and are good to go.  This full data set is available [here](https://github.com/jaytimm/uspoliticalextras/blob/master/clean-data-sets/modeling-roll-call-voting-behavior.csv).  

```{r message=FALSE, warning=FALSE}
full <- reps %>% left_join(dems) %>% left_join(nominates) %>%
  mutate(ethnicity = as.factor(ethnicity),
         gender = as.factor(gender),
         party = as.factor(party),
         is_south = as.factor(is_south))
```



## Modeling political ideology in the 116th House

```{r}
keeps <- c('nominate.dim1', 'Per_BachelorsHigher',
           'Median_HH_Income', 'Per_Black',
           'Per_Hispanic', 'is_south', 'party',
           'gender', 'ethnicity', 'age') 

full1 <- full[, c(keeps)]
colnames(full1) <- meta$var

full1 <- within(full1, member_ETHNICITY <- relevel(member_ETHNICITY, ref = 4))
full1 <- within(full1, member_GENDER <- relevel(member_GENDER, ref = 2))
full1 <- within(full1, member_PARTY <- relevel(member_PARTY, ref = 2))
full1 <- within(full1, cd_IS_SOUTH <- relevel(cd_IS_SOUTH, ref = 2))
```

```{r eval=FALSE, include=FALSE}
setwd('/home/jtimm/jt_work/GitHub/packages/uspoliticalextras/clean-data-sets')
write_csv(cbind(full[,c(1:4,6)], full1), 'modeling-roll-call-voting-behavior.csv')
```



### +Three models

Per @mccarty2016polarized, and largely for good measure here, we investigate the utility of three models in accounting for variation in DW-NOMINATE scores: (1) only constituent demographics, (2) constituent demographics and house member party, and (3) constituent demographics, house member party, and house member characteristics.

```{r}
modA <- lm(dw_NOMINATE_DIM_1 ~ 
             cd_BACHELORS + 
             cd_HH_INCOME + 
             cd_BLACK +
             cd_HISPANIC + 
             cd_IS_SOUTH, 
           data = full1)

####
modB <- lm(dw_NOMINATE_DIM_1 ~ 
             cd_BACHELORS + 
             cd_HH_INCOME + 
             cd_BLACK +
             cd_HISPANIC + 
             cd_IS_SOUTH + 
             
             member_PARTY, 
           data = full1)

####
modC <- lm(dw_NOMINATE_DIM_1 ~ 
             cd_BACHELORS + 
             cd_HH_INCOME + 
             cd_BLACK +
             cd_HISPANIC + 
             cd_IS_SOUTH +
             
             member_PARTY +
             
             member_GENDER +
             
             member_ETHNICITY +
             member_AGE, 
           data = full1)
```



### +Adjusted r-squared per model

Values are similar to those presented @mccarty2016polarized.   

```{r}
data.frame(modA = round(summary(modA)$adj.r.squared, 3),
           modB = round(summary(modB)$adj.r.squared, 3),
           modC = round(summary(modC)$adj.r.squared, 3)) %>%
  knitr::kable()
```



### +Coefficients: full model

```{r render = 'normal_print'}
td <- broom::tidy(modC) %>%
  mutate_if(is.numeric, round, 3) 

colors <- which(td$`p.value` < .05)

td %>%
  knitr::kable(booktabs = T, format = "html") %>%
  kableExtra::kable_styling() %>%
  kableExtra::row_spec(colors,  
                       background = "#e4eef4") #bold = T, color = "white",
```


### +A visual summary 

```{r fig.height=5.5}
jtools::plot_summs(modC, scale = TRUE)
```


### +Some interpretations

**In terms of constituency demographics**, then, house members representing districts with higher percentages of college grads and Hispanics tend to have lower NOMINATE scores, ie, are more liberal in voting behavior.  Also, house members representing non-Southern districts have lower scores.  

**In terms of member characteristics**, Black house members, female house members, and older house members all have lower scores as well.  Party affiliation is the strongest predictor -- simply getting elected as a Democrat amounts to a 0.776 *decrease* in NOMINATE scores (on average) relative to a Republican (again, on a scale from -1 to +1).  

For the average present day American, model results are in no way surprising.   However, as @mccarty2016polarized demonstrate (see Chapter 2 appendix), constituency demographics & member party have become increasingly more predictive of roll call voting behavior since the early 70s.  As voting behavior in the house has  become more extreme & ideologically divided.



## A final thought

So, this post has focused largely on aggregating pieces of a model puzzle as presented in @mccarty2016polarized. We have assumed quite a bit of knowledge wrt the NOMINATE research paradigm, without contextualizing or motivating model composition or results in any real way.    Read the reference for this -- it tells the story of **increasing polarization in American politics** over the last 40 years or so.  A story that becomes more relevant by the day.   


## References

